---
title: "Lab 2-2 Solution"
author: "Kun Zhang"
date: "5 December 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Earning Data
```{r initialize}
data <- read.csv("Data/finance/EconomicDataSummerSchool.csv", header=TRUE)
data2 <- read.csv("Data/finance/BoujeeCapital.csv", header=TRUE)

library(arm)
library(ggplot2)
```

## Exploratory Data Analysis
### BHP Share Price vs AUD/USD Currency
```{r BHP Share Price vs AUD/USD Currency}
x1 = data$AUD.USD
x2 = data$BHP.Share.Price
plot(x1, x2, xlab="AUD:USD", ylab="BHP Share Price")
abline(lm(x2~x1))
title("Effect of currency vs BHP Share Price")
```

### BHP Share Price vs Inflation
```{r BHP Share Price vs Inflation}
x3 = data$Inflation
x4 = data$BHP.Share.Price
plot(x3, x4, xlab="Inflation", ylab="BHP Share Price")
title("Effect of Australian GDP Growth vs BHP Share Price")
```

### BHP Share Price vs AUD:CNY Currency
```{r BHP Share Price vs AUD:CNY Currency}
x5 = data$AUD.CNY.Growth
x6 = data$BHP.Share.Price
plot(x5, x6, xlab="AUD:CNY", ylab="BHP Share Price")
title("AUD/CNY vs BHP Share Price")
```

### Distribution of BHP Share Price
```{r Distribution of BHP Share Price}
plot(density(x1), main="Distribution of AUD:USD Share Price", ylab="Frequency", xlab = "Currency")
plot(density(x2), main="Distribution of BHP Share Price", ylab="Frequency", xlab = "Share Price")
plot(density(x3), main="Australian Inflation Rate", ylab="Frequency", xlab = "Inflation Rate")
plot(density(x5), main="Distribution of AUD:CNY Growth", ylab="Frequency", xlab = "AUD:CNY")
```

## Regression Analysis 1: Fitting a simple model to these data
```{r Regression Analysis 1: Fitting a simple model to these data}
model1 <- lm(x2 ~ x1)
display(model1)
plot(model1)
```
## Regression Analysis 2: BHP Share Price
```{r Regression Analysis 2: BHP Share Price}
y = data$BHP.Share.Price
x1 = data$Unemployment.Rate
x2 = data$AUD.USD
x3 = data$Crude.oil..average
x4 = data$Iron.ore..cfr.spot
x5 = data$Gold
model2 <- lm(y ~ x1 + x2 + x3 + x4 + x5)
display(model2)
plot(model2)
```

## Regression Analysis 3: Hedge Fund Returns
Alpha:0.01
  
Beta: 0.33
  
A beta of less than 1 means that the security will be less volatile than the index.
  
The hedge fund outperform index by 1% and 66% less volatile than the index. Given the management expense ratio (MER) is usually high for hedge funds, if the return data is recorded before fee deduction, this hedge fund has very poor performance.
  
```{r Regression Analysis 3: Hedge Fund Returns}
index_returns <- data2$Managed.Futures
hf_returns <- data2$Boujee.Capital
plot(density(index_returns), main="Index Returns Histogram", ylab="Frequency", xlab = "Index Returns")
plot(density(hf_returns), main="Hedge Fund Returns Histogram", ylab="Frequency", xlab = "Hedge Fund Returns")

model3 <- lm(hf_returns ~ index_returns)
display(model3)
plot(model3) #alpha 0.01, beta 0.33
plot(index_returns, hf_returns)
abline(model3)
title("Analysis of hedge fund returns")
```









