---
title: "Lab 1+ Solution Part3"
author: "Kun Zhang"
date: "5 December 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Smoking Data Reshaping
```{r initialize}
library(plyr)
smoke.data <- read.csv("./Data/Smoking/smoke_data.csv", head=TRUE, sep=",", na.strings=" ") #na.strings = " " blank

```

## Reshaping
Note that smoke.data3 has more data than smoke.data because it incorporates NA data that are not present in smoke.data. For example, a person may only have participated in wave 3. In smoke.data, the wave 3 data is not presented, but it is presented in smoke.data as smoke.data2 adds NA values. The total smkreg data in smoke.data2 size is 1760*6 = 10560, larger than 8730 in smoke.data.
```{r Reshaping}
library(reshape2)
(smoke.data)
#dcast: expand wave; ...spread operator
(smoke.data2 <- dcast(smoke.data, ...~ wave, value.var="smkreg", na.rm=TRUE))
#melt: shrink wave
(smoke.data3 <-melt(smoke.data2, id.vars = c("newid","sex.1.F.","parsmk"), measure.vars = c("1", "2", "3", "4", "5", "6")))
```



